Cost- and Deadline-constrained Scheduling of Scientific Workflow Ensembles in IaaS Clouds

### 2012-01-20 paper/paper.tex
Some minor updates to the introduction

Scientific workflows, usually represented as directed acyclic graphs (DAG), are
the important class of applications that have been studied in the context of
resource management and scheduling on grid and utility computing systems.
However, large computational applications are not just individual workflows but
rather sets of inter-related workflows grouped in {\em ensembles}. All the
workflows in an ensemble typically have a similar structure, but they differ by
input data, number of tasks and individual tasks sizes. A good example of
scientific workflow ensemble comes from the CyberShake
application~\cite{Callaghan11}, which calculates seismic hazards for a given
geographical region such as California. In order to produce a hazard map, the
hazard curves need to be computed for a set of geographical sites, while each
site requires running a single complex scientific workflow. Workflows for each
site may differ not only by their parameters, but also by priority; e.g. some of
the points on the map correspond to urban areas or strategic locations such as
power plants, whereas others may be located in low populated regions. Another
examples of workflow ensembles come from astronomical applications, such as
Montage~\cite{Deelman08} or search for Earth-like planets from Kepler project
data~\cite{vockler11}, where multiple workflows are typically required to
process data covering different parts of the sky.

Infrastructure-as-a-service (Iaas) clouds offer capabilities of creating
(provisioning) computational resources on demand in a pay-per-use model and are regarded by
scientific community as a potentially attractive source of computing
resources~\cite{Ostermann10,Keahey09}. In contrast to clusters and grids which
typically offer best-effort resource provisioning capabilities, clouds give more
flexibility in terms of creating a controlled and managed computing environment
with the ability of adjusting resource capacity to the computing demands, often
called auto-scaling. Giving the users more control, however, clouds also require
developing new methods of task scheduling and resource provisioning. The
resource management decisions in cloud scenarios not only have to take into
account performance related metrics such as workflow makespan or resource
utilization, but also budget considerations, since the resources from public
(commercial) cloud providers are usually not free~\cite{Durkee10}.

In this paper, to get insight into these resource management challenges for
scietific workflow on clouds, we address a new important problem of maximizing
the number of completed workflows from an ensemble under both budget and
deadline constraints. The motivation is to answer the fundamental question of a
researcher: how much work can be completed in a limited budget and timeframe of
a research project. The goals of this paper are to discuss and assess possible
static and dynamic strategies for both task scheduling and resource
provisioning. Based on the knowledge of workflow scheduling algorithms we
analyze strategies for on-line scheduling of individual tasks as well as static
algorithms that rely on the information about the workflow structure (critical
paths and workflow levels) and estimates of task runtimes. In addition, we
analyze a hybrid workflow-aware dynamic scheduling algorithm, which uses the
workflow structure information to estimate which workflows should be rejected
from the ensemble due to the constraints. As methodology in the study of the
proposed algorithms we use simulation techniques. We have developed cloud workflow simulator based on
CloudSim~\cite{Calheiros11}, which models the infrastructure and the application. The
algorithms are subject to evaluation on a set of scientific workflow ensembles with a broad range of
budget and deadline parameters.

@@

\begin{algorithm}
\caption{Dynamic provisioning algorithm for DPDS}
\label{alg:prov}
\begin{algorithmic}[1]
\Require $c$: consumed budget; $b$: total budget; $d$: deadline; $p$: price;
$t$: current time; $u_h$: upper utilization threshold; $u_l$: lower utilization
threshold; $v_{max}$: maximum number of VMs
\Procedure{Provision}{}
	\State $V_R\gets$ set of running VMs
    \State $V_C\gets$ set of VMs completing billing cycle
    \If{$ b - c < |V_C| * p\ or\ t > d $ }
    	\State $n_T\gets |V_R| - \lfloor(b-c)/p\rfloor$
    	\State $V_T\gets$ select $n_T$ VMs to terminate from $V_C$
    	\State \Call{Terminate}{$V_T$} \label{l:terminate1}
    \Else 
		\State $u\gets$ current VM utilization
    	\If{$u>u_h$ and $|V_R| < v_{max}$}
    		\State \Call{Start}{$new\ VM$}
    	\ElsIf{$u<u_l$}
    		\State $V_I\gets$ set of idle VMs
    		\State $n_T\gets \lceil|V_I|/2\rceil$ \label{l:nT2}
			\State $V_T\gets$ select $n_T$ VMs to terminate from $V_I$
    		\State \Call{Terminate}{$V_T$} \label{l:terminate2}
    	\EndIf 
    \EndIf
\EndProcedure
\end{algorithmic} 
\end{algorithm}

The set of VMs completing their billing cycle is determined by considering both the
provisioner interval, and the termination delay of the provider. This guarantees 
that VMs can be terminated before they start the next billing cycle and prevents the
budget from being overrun. The VMs terminated in line \ref{l:terminate1} are the ones 
that would overrun the budget if not terminated in the current provisioning cycle. 
The VMs terminated on line \ref{l:nT2} are chosen to increase the resource utilization
to the desired threshold. In order to prevent instances from being
terminated too quickly, potentially wasting resources that have already been paid for
but could be used later, no more than half of the idle resources are terminated during
each provisioning cycle. To avoid an uncontrolled increase in the number of
instances, which may happen in the case of highly parallel workflows, the provisioner 
will not start a new VM if the user-defined scaling parameter, $v_{max}$, has
been reached.

@@

To evaluate our algorithms, we developed a cloud workflows simulator based on
CloudSim~\cite{Calheiros11}. To represent our infrastructure model, we defined
Cloud, VM and WorkflowEngine simulation entities. Cloud object is responsible
for staring and terminating VMs using API similar to Amazon EC2. VM object is
responsible for execution of individual tasks, while WorkflowEngine deals with
management of ensembles and workflows. WorkflowEngine uses separate, although
tightly coupled Scheduler and Provisioner modules. We assume that a single core
of a VM executed submitted tasks sequentially and that a task is given exclusive
access to allocated VM. Although CloudSim provides a more advanced
infrastructure model which includes time-sharing and space-sharing policies and
power consumption in a datacenter, we do not use these features, since we are
interested mainly in execution of tasks on VMs and high-level worklow scheduling
on top of that. The simulator relies on DAG and DAX format used by Pegasus
project and workflow gallery.

@@

In order to evaluate the algorithms on a standard ser of workflows, we used
ensembles constructed from the DAGs available\footnote{\url{https://confluence.pegasus.isi.edu/display/pegasus/WorkflowGenerator}}
in the workflow generator gallery~\cite{Bharathi08}. The gallery contains
workflows of five applications from Pegasus and from Indiana
University~\cite{Ramakrishnan08}. We constructed ensembles of 100 DAGs by sampling the
workflows of different sizes based on a given distribution. By the size of the
workflow means the number of jobs. Two distributions are considered:
\begin{itemize}
  \item ensemble of workflows of equal size (constant)
  \item Pareto-like distribution.
\end{itemize}

@@

For each application type, we selected such ranges of constraints (deadline,
budget) that it is possible to observe the possibly broad result space: from the
tight constraints, when only a small number of workflows can finish to the
liberal constraints when all or almost all the workflows can finish. These
parameters vary among the applications. Montage and CyberShake benchmarks
workflows required the budget in \$20 -- \$150 range and deadlines from 1 to
20 hours to complete the ensemble. LIGO and SIPHT workflows have more long
tasks, so they require budgets up to \$1,500 and deadlines extended to 40
hours. Epigenomics workflow is even more costly and needs approximately \$12,000
and 1500 hours to complete the full ensemble.

We show the results of three algorithms (TODO: ADD MORE): DPDS, WA-DPDS and
SPSS. These tests were run with maximum autoscaling factor set to 1.0 for
DPDS and WA-DPDS. After experimenting with DPDS and WA-DPDS we found that due to
high parallelism of workflows the resource utilization remains high most of the time, 
which leads to rapid provisioning of new resources at the beginning of ensemble
execution. This in turn results in less efficient total utilization due to high
level of parallelism. Setting the scaling limit to 1.0, ensures that the shape
of resource provisioning plan is closest to a rectangle (see
Fig.~\ref{fig:spds-example}) and the VM termination procedure
(Algorithm~\ref{alg:prov}) allows maximum utilization of VMs before shutdown.

>>>>

Scientific workflows, usually represented as directed acyclic graphs (DAG), are
the important class of applications that have been studied in the context of
resource management and scheduling on grid and utility computing systems.
However, large computational applications are often composed of several 
inter-related workflows grouped into {\em ensembles}. All the
workflows in an ensemble typically have a similar structure, but they differ by
input data, number of tasks and individual tasks sizes. A good example of
scientific workflow ensemble comes from the CyberShake
application~\cite{Callaghan11}, which calculates seismic hazards for a given
geographical region such as California. In order to produce a hazard map, the
hazard curves need to be computed for a set of geographical sites, while each
site requires running a single complex scientific workflow. Workflows for each
site may differ not only by their parameters, but also by priority; e.g. some of
the points on the map correspond to urban areas or strategic locations such as
power plants, whereas others may be located in less populated regions. Another
example of workflow ensembles comes from astronomical applications, such as
Montage~\cite{Deelman08} or search for Earth-like planets from Kepler project
data~\cite{vockler11}, where multiple, different workflows are used to
process data for different areas of the sky.

Infrastructure-as-a-Service (IaaS) clouds offer the ability to provision 
computational resources on-demand according to a pay-per-use model. These systems 
are regarded by the scientific community as a potentially attractive source of 
low-cost, on-demand computing resources~\cite{Ostermann10,Keahey09}. In contrast 
to clusters and grids, which typically offer best-effort quality of service, clouds 
give users more flexibility to creating a controlled and managed computing environment
with the ability to adjust resource capacity according to the dynamically changing
computing demands of the application, often called auto-scaling. Giving the users 
more control, however, also requires the development of new methods for task 
scheduling and resource provisioning. The resource management decisions required 
in cloud scenarios not only have to take into account traditional performance-related 
metrics such as workflow makespan or resource utilization, but must also now consider
budget constraints, since the resources from public (commercial) cloud providers 
are usually not free~\cite{Durkee10}.

In this paper, to get insight into these resource management challenges for
scientific workflows on clouds, we address a new important problem of maximizing
the number of completed workflows from an ensemble under both budget and
deadline constraints. The motivation is to answer the fundamental question of a
researcher: How much computation can be completed given the limited budget and 
timeframe of a research project? The goals of this paper are to discuss and assess possible
static and dynamic strategies for both task scheduling and resource
provisioning. Based on the knowledge of workflow scheduling algorithms we
analyze strategies for on-line scheduling of individual tasks as well as static
algorithms that rely on the information about the workflow structure (critical
paths and workflow levels) and estimates of task runtimes. In addition, we
analyze a hybrid workflow-aware dynamic scheduling algorithm, which uses the
workflow structure information to estimate which workflows should be rejected
from the ensemble due to the constraints. As methodology in the study of the
proposed algorithms we use simulation techniques. We have developed cloud workflow simulator based on
CloudSim~\cite{Calheiros11}, which models the infrastructure and the application. The
algorithms are subject to evaluation on a set of scientific workflow ensembles with a broad range of
budget and deadline parameters.

@@

\begin{algorithm}
\caption{Dynamic provisioning algorithm for DPDS}
\label{alg:prov}
\begin{algorithmic}[1]
\Require $c$: consumed budget; $b$: total budget; $d$: deadline; $p$: price;
$t$: current time; $u_h$: upper utilization threshold; $u_l$: lower utilization
threshold; $v_{max}$: maximum number of VMs
\Procedure{Provision}{}
	\State $V_R\gets$ set of running VMs
    \State $V_C\gets$ set of VMs completing billing cycle
    \If{$ b - c < |V_C| * p\ or\ t > d $ }
    	\State $n_T\gets |V_R| - \lfloor(b-c)/p\rfloor$
    	\State $V_T\gets$ select $n_T$ VMs to terminate from $V_C$
    	\State \Call{Terminate}{$V_T$} \label{l:terminate1}
    \Else 
		\State $u\gets$ current VM utilization
    	\If{$u>u_h$ and $|V_R| < v_{max}*N_{VM}$}
    		\State \Call{Start}{$new\ VM$}
    	\ElsIf{$u<u_l$}
    		\State $V_I\gets$ set of idle VMs
    		\State $n_T\gets \lceil|V_I|/2\rceil$ \label{l:nT2}
			\State $V_T\gets$ select $n_T$ VMs to terminate from $V_I$
    		\State \Call{Terminate}{$V_T$} \label{l:terminate2}
    	\EndIf 
    \EndIf
\EndProcedure
\end{algorithmic} 
\end{algorithm}

The set of VMs completing their billing cycle is determined by considering both the
provisioner interval, and the termination delay of the provider. This guarantees 
that VMs can be terminated before they start the next billing cycle and prevents the
budget from being overrun. The VMs terminated in line \ref{l:terminate1} are the ones 
that would overrun the budget if not terminated in the current provisioning cycle. 
The VMs terminated on line \ref{l:nT2} are chosen to increase the resource utilization
to the desired threshold. In order to prevent instances from being
terminated too quickly, potentially wasting resources that have already been paid for
but could be used later, no more than half of the idle resources are terminated during
each provisioning cycle. To avoid an uncontrolled increase in the number of
instances, which may happen in the case of highly parallel workflows, the provisioner 
will not start a new VM if the number of running VMs is greater than the product of
$N_{VM}$ and an autoscaling factor, $v_{max}$.

@@

To evaluate our algorithms, we developed a cloud workflows simulator based on
CloudSim~\cite{Calheiros11}. To represent our infrastructure model, we defined
Cloud, VM and WorkflowEngine simulation entities. Cloud object is responsible
for staring and terminating VMs using API similar to Amazon EC2. VM object is
responsible for execution of individual tasks, while WorkflowEngine deals with
management of ensembles and workflows. WorkflowEngine uses separate, although
tightly coupled Scheduler and Provisioner modules. We assume that a single core
of a VM executed submitted tasks sequentially and that a task is given exclusive
access to allocated VM. Although CloudSim provides a more advanced
infrastructure model which includes time-sharing and space-sharing policies and
power consumption in a datacenter, we do not use these features, since we are
interested mainly in execution of tasks on VMs and high-level workflow scheduling
on top of that. The simulator relies on DAG and DAX format used by Pegasus
project and workflow gallery.

@@

In order to evaluate the algorithms on a standard set of workflows, we used
ensembles constructed from the DAGs available\footnote{\url{https://confluence.pegasus.isi.edu/display/pegasus/WorkflowGenerator}}
in the workflow generator gallery~\cite{Bharathi08}. The gallery contains
workflows of five applications from Pegasus as well as several other applications 
from~\cite{Ramakrishnan08}. We constructed ensembles of 100 DAGs by sampling the
workflows of different sizes based on a given distribution. By the size of the
workflow means the number of jobs. Two distributions are considered:
\begin{itemize}
  \item ensemble of workflows of equal size (constant)
  \item Pareto-like distribution.
\end{itemize}

@@

For each application type, we selected such ranges of constraints (deadline,
budget) that it is possible to observe the possibly broad result space: from the
tight constraints, when only a small number of workflows can finish to the
liberal constraints when all or almost all the workflows can finish. These
parameters vary among the applications. Montage and CyberShake benchmarks
workflows required the budget in \$20 -- \$150 range and deadlines from 1 to
20 hours to complete the ensemble. LIGO and SIPHT workflows have more long
tasks, so they require budgets up to \$1,500 and deadlines extended to 40
hours. The Epigenomics workflow is even more costly and needs approximately \$12,000
and 1500 hours to complete the full ensemble.

We show the results for three algorithms: DPDS, WA-DPDS and
SPSS. These tests were run with maximum autoscaling factor set to 1.0 for
DPDS and WA-DPDS. After experimenting with DPDS and WA-DPDS we found that due to
high parallelism of workflows the resource utilization remains high most of the time, 
which leads to rapid provisioning of new resources at the beginning of ensemble
execution. This in turn results in less efficient total utilization due to high
level of parallelism. Setting the scaling limit to 1.0, ensures that the shape
of resource provisioning plan is closest to a rectangle (see
Fig.~\ref{fig:spds-example}) and the VM termination procedure
(Algorithm~\ref{alg:prov}) allows maximum utilization of VMs before shutdown.
